Penalized regression for left-truncated and right-censored survival data

### 2021-01-05 manuscript.tex
new edits

\begin{abstract}
    High-dimensional data are common in the medical field as high-throughput screening, detailed electronic health record (EHR) information, and comprehensive genomics testing are now used to capture the patient profile. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

@@

\section{Introduction}
The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information. The level of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes. Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}. Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability. Survival analysis is of particular interest given that overall survival is often a primary outcome of interest.

In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model. Similar to ordinary least squares regression (OLS), these penalized regression methods estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients. This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection. Common penalized approaches include lasso \citep{tibshirani1996regression}, ridge \citep{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \citep{zou2005regularization}, with extensions such as the adaptive lasso \citep{zou2006adaptive}.

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data so their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times.

@@

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time and leave only at death or censoring. However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\caption{Patient follow-up. Left-truncated, right-censored patient follow-up in a hypothetical study cohort. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time) and the recently-diagnosed.

Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time. Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$. The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$. In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (e.g. exceeded their left truncation time). Here, event times $t_1 < t_2 < ... < t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort, rather than assuming that all patients are at risk starting at the same referent time $0$.

@@

We take a subset of $\nPatients$ stage IV, non-small cell lung cancer (NSCLC) patients in the CGDB as a use case for this study. The outcome of interest is survival time predicted from the date of Stage 4 diagnosis. Survival times are left-truncated because patients who die before receiving a Foundation Medicine test are logically not included in the cohort of Foundation Medicine test recipients.

As shown in \autoref{fig:left-trunc-time} the distribution of left truncation times (time from diagnosis to genomic testing) is right-tailed with approximately $\entryMoreOneYear$ of patients diagnosed more than one year prior to receiving the Foundation Medicine genomic test.  Moreover, the correlation between diagnosis year and left truncation time is strongly negative ({$\rho$} = $\corrDxYear$), suggesting the presence of left truncation survivor bias as patients with records in the CGDB diagnosed longer ago needed to survive more years before being tested.

@@

This process was repeated $20$ times. For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients. Simulations were peformed in both big and small $p$ scenarios. In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the big $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

@@

The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line on both extremes. Predictions tend to be more accurate for patients with predicted survival close to $0.5$, but underestimated for patients with higher survival probabilities and overestimated for patients with lower survival probabilities. Probability rescaling has been suggested in such cases (e.g. characteristic sigmoid-shaped calibration curves produced by different learning algorithms), for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

@@

These results are perhaps not surprising. For example, when computing the C-index in the observed sample the model is evaluated in a test set that suffers from the same bias as the training set. Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry. In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated. As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients can be considerably larger if there is a more direct correlation.

@@

The larger model included all non-biomarker variables from the smaller model and added all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data only to columns with less than 50\% missing indicators and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio above 2 in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.

@@

\autoref{fig:surv-calibration} displays the same survival calibration curves presented in the simulation section. As in the simulation study, the unadjusted models overpredict survival probabilities. In contrast, the adjusted model tends to underpredict survival, which contrasts from the well calibrated adjusted models from the simulations. Here, the adjusted models tend to be adequately calibrated for patients with lower survival probabilities but less well calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients.

@@

\section{Discussion} \label{sec:discussion}
In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically assess its impact on bias and inference in left truncated data. Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data. In particular, our simulation shows that predictions from models that fail to adjust for left truncation will overestimate true survival probabilities whereas models that properly adjust can yield well calibrated survival predictions, even with high-dimensional data. Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation.

Our results have important implications for specification, interpretation, and evaluation of prognostic survival models. Much of the difficulty arises because training and test sets are both subject to the same selection bias. In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray. For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not and may therefore lead to misleading conclusions in the presence of left truncated data. On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.

@@

We conclude with several recommendations. When making predictions with survival data is of interest, it is critical to define the risk set appropriately. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the database of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking this ``delayed entry'' of at-risk patients would lead to adverse consequences including misleading conclusions, as we demonstrate in this work. In addition, we recommend that models should be evaluated using both discrimination and calibration based metrics.

>>>>

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

@@

\section{Introduction}
The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), clinical (electronic health records), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information. The amount of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes. Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}. Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability. Survival analysis is of particular interest given that overall survival is often a primary outcome of interest.

In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model. Similar to ordinary least squares (OLS) regression, these penalized models estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients. This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection. Common penalized approaches include lasso \citep{tibshirani1996regression}, ridge \citep{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \citep{zou2005regularization}, with extensions such as the adaptive lasso \citep{zou2006adaptive}.

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times. Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}. These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.

@@

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring. However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\caption{Left-truncated, right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.

Fortunately, the partial likelihood can be easily modified to make $R_i$ the set of individuals who are at risk (have entered the study and are not censored or failed) at that time. Let individual $i$ have survival data of the form $(v_i, y_i, e_i, x_i)$, where individual $i$ enters the study at time $v_i$. The left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$. In other words, subjects are only considered at risk in the analysis after they have been observed to be at risk (i.e. exceeded their left truncation time). Here, event times $t_1 < t_2 < ... < t_m$ are ordered relative to the earliest chronological event time observed in the patient cohort (as shown in \autoref{fig:followup}), rather than assuming that all patients are at risk starting at the same referent time $0$.

@@

We take a subset of $\nPatients$ stage IV, non-small cell lung cancer (NSCLC) patients in the CGDB as a use case for this study. The outcome of interest is survival time predicted from the date of stage IV diagnosis. Survival times are left-truncated because patients who die before receiving a Foundation Medicine test are logically not included in the cohort of Foundation Medicine test recipients.

As shown in \autoref{fig:left-trunc-time}, the distribution of left truncation times (time from diagnosis to genomic testing) is right-tailed with approximately $\entryMoreOneYear$ of patients diagnosed more than one year prior to receiving the Foundation Medicine genomic test.  Moreover, the correlation between diagnosis year and left truncation time is strongly negative ({$\rho$} = $\corrDxYear$), suggesting the presence of left truncation survivor bias as patients with records in the CGDB diagnosed longer ago needed to survive more years before the availability of genomic testing.

@@

This process was repeated $200$ times. For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients. Simulations were peformed in both big and small $p$ scenarios. In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the big $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

@@

The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities. Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but underestimated for patients with higher survival probabilities. Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

@@

These results are perhaps not surprising. For example, when computing the C-index in the observed sample, the model is evaluated in a test set that suffers from the same bias as the training set. Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry. In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated. As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients can be considerably larger if there is a more direct correlation.

@@

The larger model included all non-biomarker variables from the smaller model and added all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data only to biomarkers with less than 50\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.

@@

\autoref{fig:surv-calibration} displays the same survival calibration curves presented in the simulation section. As in the simulation study, the unadjusted models overpredict survival probabilities. In contrast, the adjusted model tends to underpredict survival, which contrasts from the well calibrated adjusted models from the simulations. Here, the adjusted models tend to be adequately calibrated for patients with lower survival probabilities but less well calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and do not adequately separate low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso.

@@

\section{Discussion} \label{sec:discussion}
In this paper, we implement an efficient approach for a left truncation-adjusted, penalized Cox proportional hazards model and critically examine its impact on bias and inference in left truncated data. Using simulation studies and examples from real-world EHR and genomics data, we showed that there is a need for approaches that can both adjust for left truncation and model high dimensional data. In particular, our simulation shows that predictions from models that fail to adjust for left truncation will overestimate true survival probabilities whereas models that properly adjust can yield well calibrated survival predictions, even with high-dimensional data. Moreover, in our real-world use case, there were significant differences in coefficient estimates between models that did and did not adjust for left truncation, compromising inference.

Our results have important implications for specification, interpretation, and evaluation of prognostic survival models. Much of the difficulty arises because training and test sets are both subject to the same selection bias. In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray. For example, our simulations suggest that commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of left truncated data. On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.

@@

We conclude with several recommendations. When making predictions with survival data is of interest, it is critical to define the risk set appropriately. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the database of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking this ``delayed entry'' of at-risk patients would lead to adverse consequences including misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is not quasi-independent of survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, there exist formal tests to assess quasi-independence via the conditional Kendall's tau \citep{martin2005tau}. Finally, we recommend that models should be evaluated using both discrimination and calibration based metrics.
